A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-15 tex/sec5-synthesis.tex
[ tex ] Minor

\LT{What induction principle are we using?}

@@

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $t$ alone).
  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
  The last case \Rule{Op} is the interesting part of this theorem.
  \begin{description}
    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |-_{\Sigma, \Omega} x \syn A$.
      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |-_{\Sigma, \Omega} x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
    \item[\SynRule{Anno}:] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma |- t \chk A$ is derivable or not by induction hypothesis.
      \begin{enumerate}
        \item If $\Gamma |- t \chk A$ is derivable, then $\Gamma |- (t : A) \syn A$ is derivable.
        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- (t : A) \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
      \end{enumerate}

    \item[\ChkRule{Sub}:] If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{enumerate}
        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
          \begin{enumerate}
            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t\subsum \chk A$.
            \item if $A \neq B$ but $\Gamma |- t \subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
          \end{enumerate}
        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t\subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
      \end{enumerate}

%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
%      \]
    \item[\Rule{Op}:]
      Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.
      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub(\Xi, \emptyset)$ incrementally which is defined nowhere initially if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.

>>>>

@@

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $t$ alone).
  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
  The last case \Rule{Op} is the interesting part of this theorem.
  \begin{description}
    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- x \syn A$.
      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |- x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
    \item[\SynRule{Anno}:] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma |- t \chk A$ is derivable or not by induction hypothesis.
      \begin{enumerate}
        \item If $\Gamma |- t \chk A$ is derivable, then $\Gamma |- (t : A) \syn A$ is derivable.
        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- (t : A) \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
      \end{enumerate}

    \item[\ChkRule{Sub}:] If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{enumerate}
        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
          \begin{enumerate}
            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
          \end{enumerate}
        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
      \end{enumerate}

%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
%      \]
    \item[\Rule{Op}:]
      Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.
      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.
